Many Analystis Project

### 2021-10-01 analysis.Rmd
Update analysis.Rmd

```{r}
# get data into shape
d1 <- df %>%
  dplyr::filter(
    !is.na(wb_psychANXIETY),
    !is.na(age),
    !is.na(rel_CHURCHFREQ),
    !is.na(wb_psychMEANING),
    !is.na(male),
    !is.na(rel_EXTENTBELIEVEGOD),
  ) %>%
  dplyr::mutate(
    anx = as.integer(6- wb_psychANXIETY), # reverse coded! rescale 1-5
    anx_s = scale(wb_psychANXIETY),
    #put outcome in ST units (see McElreath 2021)  # not necessary
    age_s = scale(age),
    ch = rel_CHURCHFREQ,
    mn  = wb_psychMEANING,
    bg =  rel_EXTENTBELIEVEGOD,
    ch_s = scale(rel_CHURCHFREQ),
    mn_s = scale(wb_psychMEANING),
    bg_s = scale(rel_EXTENTBELIEVEGOD)
  ) %>%
  dplyr::select(anx,
                ch,
                mn,
                bg,
                anx_s,
                age_s,
                ch_s,
                mn_s,
                male2,
                country,
                male,
                bg_s,
                wb_psychANXIETY)

@@

@@

```{r echo}
glimpse(d1)
m1o <- brm(
    anx ~ ch_s +
    age_s  +
    male +
    (1 + ch_s | country),
  data = d1,
  family = cumulative(link ="logit"),
  seed = 1234,
  warmup = 1000,
  iter = 3000,
  chains = 4,
  file = here::here("models", "m1o_real")
)
```

@@

```{r}
parameters::model_parameters(m1o, test = "p_direction")
parameters::random_parameters(m1o, component = "conditional")
```

@@

```{r}
parF<-parameters::model_parameters(m1o ) 
plot(parF,  sort = TRUE) +theme_lucid() #option -->
```

@@

```{r}
parR<-parameters::model_parameters(m1o,effects = "random" ) 
plot(parR,  sort = TRUE) +theme_lucid() #option -->
```

@@

```{r}
plot(ggeffects::ggpredict(
  m1o,
  terms = c("country[all]", "ch_s[minmax]"),
  type = "re",
  ci.lvl = 0.89
)) + scale_color_viridis_d() + 
  theme_classic()
```

@@

```{r, cache = TRUE}
library(viridis)
p0b<-brms::conditional_effects(m1o, "ch_s", 
                              spaghetti = T,
                              ndraws = 200,
                              categorical = T,
                              prob = .89,
                              method = 'fitted')

@@

@@

### Model
```{r}
m2o <- brm(
  anx ~ bg_s +
    age_s  +
    male +
    (1 + bg_s | country),
  data = d1,
  family = cumulative(link = "logit"),
  seed = 1234,
  warmup = 1000,
  iter = 3000,
  chains = 4,
  file = here::here("models", "m2o_real.rds")
)
```

No clear predictive effect of belief on anxiety

```{r}
parameters::model_parameters(m2o, test = "p_direction")
parameters::random_parameters(m2o)
```

```{r}
parF<-parameters::model_parameters( m2o ) 
plot(parF,  sort = TRUE) +theme_lucid() #option -->
```

```{r}
parR<-parameters::model_parameters(m2o,effects = "random" ) 
plot(parR,  sort = TRUE) +theme_lucid() #option -->
```

No country level predictions in the slope/intercept covariance:

```{r}
plot(ggeffects::ggpredict(
  m2o,
  terms = c("country[all]", "bg_s[minmax]"),
  type = "re",
  ci.lvl = 0.89
)) +
  scale_color_viridis_d() +
  theme_classic()
```

```{r, cache = TRUE}
library(viridis)
p0 <- brms::conditional_effects(
  m2o,
  "bg_s",
  spaghetti = T,
  categorical = TRUE,
  ndraws = 200,
  prob = .89,
  method = 'posterior_epred'
)

### Checks

@@

@@

There is substantial variation in country-level intercepts for anxiety. And substantial cultural level variation in religious practice and belief. All this needs to be explained. Finally India is an outlier.

@@

## Causal model

Suppose we believe that:

-   religion + ses + male + country + a different unmeasured latent variable (e.g. privilege) causes well being (citations... )

-   age + male + country + and Unmeasured latent variable (e.g. parent's religion, call this U1) causes Religion (citations... )

-   country + age + another unmeasured latent variable (e.g. privilege U2) causes ses (citations...)

-   our "exposure" is religion and our outcome is "well-being"

### Which variables should we condition on to understand the relationship between religion and well-being?

We have two sets adjustments sets to chose from. We will pick the set with the measured indicators.

```{r}
tidy_ggdag <- dagify(
  wb  ~ rel + ses + male + country +  U2,
  rel ~ age + male + country + U1,
  ses ~ country + age + U2,
  exposure =  "rel",
  outcome =   "wb"
) %>%
  tidy_dagitty()

# graph adjustment sets
ggdag::ggdag_adjustment_set(tidy_ggdag, node_size = 14) +
  theme(legend.position = "bottom") +
  theme_dag_blank()
```

### Note: there is collider counfounding if we include all variables into our model

Suppose we were to include *all* the variables in our model, except U2 and U1 which are not measured?

In this case we would introduce collider bias!

```{r}
ggdag::ggdag_collider(tidy_ggdag,
                      from = "rel",
                      to = "wb",
                      control_for(c("ses", "male", "country", "age")))

### Specifying the mechanism in relation to this dataset.

We have yet to consider a mechanism by which religion might affect well-being, which will tell us specifically which aspects of well-being and which aspects of religion to focus on. Just as it might be tempting to throw all variables into a regression equation, and disastrous to do so, it might be tempting to leap into data reduction combining all variables without pausing to think what we want to know.

There are many theories of how religion affects well-being we might consider (cite Highland et al). We will focus on two non-trivial theories -- one related to religious behavior and another related to religious belief. (Citations to follow ...)

## Church attendance

We propose that church attendance reduces psychological anxiety by increasing perceived social support (which is measured in this dataset as a well-being variable).

This is a new model so we need to write our dag all over again. The dag will resemble the dag above, except we know that people who identify as male have less social support (typically), at least in some countries.

We write the dag:

### Causal Dag: church

```{r}
tidy_ggdag2 <- dagify(
  wb  ~ support + ses + male + country + U2,
  support ~ church + male + country + den,
  church ~ age + male + country + den + U1,
  ses ~ country + age + U2,
  exposure =  "church",
  outcome =   "wb"
) %>%
  tidy_dagitty()

# graph adjustment sets
ggdag::ggdag_adjustment_set(tidy_ggdag2, node_size = 14) +
  theme(legend.position = "bottom") +
  theme_dag_blank()
```

The two adjustment sets are the same as in the original dag that we considered.

#### Confounder bias if we include our theoretically proposed causal mediators!

Before proceeding with the model, note a curiosity. If our interest is in the direct effects of church attendance on well-being we should *not* include social support because this introduces a collider bias.

```{r}
ggdag::ggdag_collider(tidy_ggdag2,
                      from = "church",
                      to = "wb",
                      controlling_for(c("male", "support", "country", "age")))

Let this serve as a warning for us to think about our causation. We must develop our causal model before we leap into our statistical model.

@@

```{r}
tidy_ggdag2 <- dagify(
  wb  ~ meaning + ses + male + country  + U2,
  meaning ~ bg +  male + country,
  bg ~ age + male + country + U1,
  ses ~ country + age + U2,
  exposure =  "bg",
  outcome =   "wb"
) %>%
  tidy_dagitty()

# graph adjustment sets
ggdag::ggdag_adjustment_set(tidy_ggdag2, node_size = 14) +
  theme(legend.position = "bottom") +
  theme_dag_blank()

tidy_ggdag2 <- dagify(
  wb  ~ meaning + ses + male + country  + U2,
  meaning ~ bg +  male + country,
  bg ~ age + male + country + U1,
  ses ~ country + age + U2,
  exposure =  "bg",
  outcome =   "wb"
) %>%
  tidy_dagitty()

# graph adjustment sets
ggdag::ggdag_adjustment_set(tidy_ggdag2, node_size = 14) +
  theme(legend.position = "bottom") +
  theme_dag_blank()

@@

```{r}
ggdag::ggdag_collider(tidy_ggdag2,
                      from = "church",
                      to = "wb",
                      controlling_for(c("male", "support", "country", "age")))
```

### Causal DAG: belief

From causal model we write our statistical model:

```{r}
m2 <- brm(
  anx_s ~
    bg_s  +
    age_s  +
    male +
    (1 + bg_s | country),
  data = d1,
  family = "gaussian",
  seed = 1234,
  warmup = 1000,
  iter = 3000,
  chains = 4,
  file = "m2_real"
)
parameters::model_parameters(m2, test = "p_direction")
brms::ranef(m2)
summary(m2)
```

>>>>

```{r}
# get data into shape
d1 <- df %>%
  dplyr::filter(
    !is.na(wb_psychANXIETY),
    !is.na(age),
    !is.na(rel_CHURCHFREQ),
   # !is.na(wb_psychMEANING),
    !is.na(male),
    !is.na(rel_EXTENTBELIEVEGOD),
  ) %>%
  dplyr::mutate(
    anx = as.integer(6- wb_psychANXIETY), # reverse coded! rescale 1-5
    anx_s = scale(wb_psychANXIETY),
    #put outcome in ST units (see McElreath 2021)  # not necessary
    age_s = scale(age),
    ch = rel_CHURCHFREQ,
    mn  = wb_psychMEANING,
    bg =  rel_EXTENTBELIEVEGOD,
    ch_s = scale(rel_CHURCHFREQ),
    mn_s = scale(wb_psychMEANING),
    bg_s = scale(rel_EXTENTBELIEVEGOD)
  ) %>%
  dplyr::select(anx,
                ch,
                mn,
                bg,
                anx_s,
                age_s,
                ch_s,
                mn_s,
                male2,
                country,
                male,
                bg_s,
                wb_psychANXIETY)

@@

## Causal Model: belief in God

```{r}
library("ggdag")
library("dagitty")
dg1a <- dagify(
  anx  ~ belief + church +  ses + male + country + U,
  belief ~ age + male + country + church + U,
  church ~ age + male + country + U,
  ses ~ country + age + edu,
  edu ~ country + male, 
  exposure =  "belief",
  outcome =   "anx",
  latent = "U",
  labels = c(
    "belief" = "belief God",
    "church" = "religious service attendance",
    "anx" = "anxiety",
    "edu" = "education",
    "ses" = "SES",
    "male" = "male",
    "country" = "country",
    "age" = "age",
    "U" = "Unmeasured
    distressing life events"
  )
) #%>%
 # tidy_dagitty()

d0a <-ggdag::ggdag(dg1a, text = FALSE, use_labels = "label")  + theme_dag() + ggtitle("What if we were to assume distressing events affect religion?")
# graph adjustment sets

d1a<-ggdag::ggdag_adjustment_set(dg1a, node_size = 14, text = FALSE, use_labels = "label") +
  theme(legend.position = "bottom") +
  theme_dag() + ggtitle("Then crossectional data could not avoid confounding")

d1a

dg1 <- dagify(
  anx  ~ belief + church +  ses + male + country + U,
  belief ~ age + male + country + church,
  church ~ age + male + country,
  ses ~ country + age + edu,
  edu ~ country + male, 
  exposure =  "belief",
  outcome =   "anx",
  latent = "U",
  labels = c(
    "belief" = "belief God",
    "church" = "religious service attendance",
    "anx" = "anxiety",
    "edu" = "education",
    "ses" = "SES",
    "male" = "male",
    "country" = "country",
    "age" = "age",
    "U" = "Unmeasured
    distressing life events"
  )
) #%>%
 # tidy_dagitty()

d0 <-ggdag::ggdag(dg1, text = FALSE, use_labels = "label")  + theme_dag() + ggtitle("Assumed Causal Relations")
# graph adjustment sets

d1<-ggdag::ggdag_adjustment_set(dg1, node_size = 14, text = FALSE, use_labels = "label") +
  theme(legend.position = "bottom") +
  theme_dag() + ggtitle("Adjustment sets for Belief in God")

d1
```

## Causal model: religious service

We condition on the following model:

-   religion + ses + male + country + a different unmeasured latent variable1 (e.g. privilege) affects distress.

-   age + male + country + and unmeasured latent variable2 (e.g. parent's religion, call this U1) affects Religion

-   country + age + an unmeasured latent variable1 causes ses (citations...)

-   our "exposure" is religious service attendance or religious belief and our outcome is "anxiety"

```{r}
library("ggdag")
dg2 <- dagify(
  anx  ~ belief + church +  ses + male + country,
  belief ~ age + male + country + church + U,
  church ~ age + male + country,
  ses ~ country + age + edu,
  edu ~ country + male, 
  exposure =  "church",
  outcome =   "anx",
  latent = "U",
  labels = c(
    "belief" = "belief God",
    "church" = "religious service attendance",
    "anx" = "anxiety",
    "edu" = "education",
    "ses" = "SES",
    "male" = "male",
    "country" = "country",
    "age" = "age",
    "U" = "Unmeasured
    distressing life events"
  )
) %>%
  tidy_dagitty()

ggdag::ggdag(dg2) + theme_dag()

ggdag::ggdag(dg2, text = FALSE, use_labels = "label")  + theme_dag()
# graph adjustment sets
d2<-ggdag::ggdag_adjustment_set(dg2, node_size = 14, text = FALSE, use_labels = "label") +
  theme(legend.position = "bottom") +
  theme_dag() + ggtitle("Adjustment set for Religious Service")

d2
```

Graph for paper

```{r}
library(patchwork)
( (d0)/(d1 + d2) +   plot_layout(guides = 'collect'))  +((d0a) + (d1a)) + plot_annotation(title = 'If distressing life events affect religious belief and behaviour, longtidinal data are needed for causal inference', tag_levels = "a")
# 14/12
```

Graph for confounding

```{r}
library(patchwork)

###  Model
```{r}
m1 <- brm(
  anx ~
    bg_s +
    ch_s +
    age_s  +
    male +
    (1 + bg_s | country),
  data = d1,
  family = cumulative(link = "logit"),
  seed = 1234,
  warmup = 1000,
  iter = 3000,
  chains = 4,
  file = here::here("models", "m2o_real")
)

parameters::model_parameters(m1, test = "p_direction")
brms::ranef(m1)
summary(m1)
```

Believe in God predicts less anxiety

```{r}
parF1 <- parameters::model_parameters(m1 ) 
plot(parF1,  sort = TRUE) +theme_lucid() #option -->
```

```{r}
parR1<-parameters::model_parameters(m1,effects = "random" ) 
plot(parR1,  sort = TRUE) +theme_lucid() #option -->
```

```{r}
plot(ggeffects::ggpredict(
  m1,
  terms = c("country[all]", "bg_s[minmax]"),
  type = "re",
  ci.lvl = 0.89
)) + scale_color_viridis_d() +  scale_y_continuous(limits = c(0,1)) + 
  theme_bw()
```

```{r, cache = TRUE}
library(viridis)
p0<-brms::conditional_effects(m1, "bg_s", 
                              spaghetti = T,
                              ndraws = 200,
                              categorical = T,
                              prob = .89,
                              method = 'fitted')

plot(p0, plot = FALSE)[[1]]   + xlab("bg_s") + ylab ("anxiety 1-5") +
  labs(title = "Predicted effects of bg_s")
```

```{r}
# YES
pp_check(m1)
```

```{r, cache = TRUE}
mcmc_plot(m1,
          type = "trace")
```

@@

```{r echo}
glimpse(d1)
m2 <- brm(
    anx ~ ch_s +
    age_s  +
    male +
    (1 + ch_s | country),
  data = d1,
  family = cumulative(link ="logit"),
  seed = 1234,
  warmup = 1000,
  iter = 3000,
  chains = 4,
  file = here::here("models", "m1o_real")
)
```

@@

@@

```{r}
parF<-parameters::model_parameters(m2 ) 
plot(parF,  sort = TRUE) +theme_lucid() #option -->
```

@@

```{r}
parR<-parameters::model_parameters(m2,effects = "random" ) 
plot(parR,  sort = TRUE) +theme_lucid() #option -->
```

@@

```{r}
plot(ggeffects::ggpredict(
  m2,
  terms = c("country[all]", "ch_s[minmax]"),
  type = "re",
  ci.lvl = 0.89
)) + scale_color_viridis_d() + 
  theme_classic()
```

@@

```{r, cache = TRUE}
library(viridis)
p0b<-brms::conditional_effects(m2, "ch_s", 
                              spaghetti = T,
                              ndraws = 200,
                              categorical = T,
                              prob = .89,
                              method = 'fitted')

@@

```{r}
# YES
pp_check(m2)
```

@@

```{r, cache = TRUE}
mcmc_plot(m2,
          type = "trace")
```

@@

Belief in God predicts anxiety. This is consistent with theories that anxiety leads to greater demand for belif in God. However, this question awaits future research.

Church attendance positively predicts lower anxiety.

@@

There is substantial variation in country-level intercepts for anxiety. And substantial cultural level variation in religious practice and belief. All this needs to be explained. India is an outlier.

@@

@@

@@

```{r}
ggdag::ggdag_collider(dg1,
                      from = "church",
                      to = "wb",
                      controlling_for(c("male", "support", "country", "age")))
```
